---
title: "Portfolio 5: Parallel Rcpp"
author: "Kieran Morris"
output:
      pdf_document:

---

For this document we will be using the Irish smart meters dataset from the `electbook` package. We first import the data and display the first few rows.

# Preprocessing

```{r}

library(electBook)
library(skimr)
data("Irish")
```


We concatonate the electricity data into one vector and center it:

```{r}
y <- do.call("c", Irish$indCons)
y <- y - mean(y)
```

and we plot it against time of day:

```{r}
ncust <- ncol(Irish$indCons)

x <- rep(Irish$extra$tod, ncust)

n <- length(x)
ss <- sample(1:n, 1e4)
plot(x[ss], y[ss], col = "grey")

```

# Simple Regression R

We will now fit a simple linear regression model to the data using the following function:

```{r}
reg1D <- function(y, x){
  
  b <- t(x) %*% y / (t(x) %*% x)  
  
  return(b)
  
}

```

This function actually performs very fast, although that's probably because it barely does any computation.

```{r}
system.time( reg1D(y, x) )[3]
```

# Simple Regression with RcppParallel

We will now implement the same function using RcppParallel. We will use inline Rcpp to do this.
```{r setup, include=FALSE}
library(Rcpp)
library(RcppParallel)
knitr::opts_chunk$set(echo = TRUE)
```

```{R}
Rcpp::sourceCpp(code='
// [[Rcpp::plugins(openmp)]]

#include <Rcpp.h>
#ifdef _OPENMP
#include <omp.h>
#endif

using namespace Rcpp;

// [[Rcpp::export]]
double parallelReg1D(NumericVector y, NumericVector x) {
   int n = x.size();
   double b = 0.0;
   double x_dot_x = 0.0;
   
   // Use OpenMP to perform the computation in parallel
   #pragma omp parallel for reduction(+:b, x_dot_x)
   for (int i = 0; i < n; i++) {
      b += x[i] * y[i];
      x_dot_x += x[i] * x[i];
   }
   
   return b / x_dot_x;
}
')
```

```{r}
parallelReg1D(y, x)
reg1D(y, x)
```

As we can see they line up perfectly! We can also check the speed of the parallel function by using microbenchmark.

```{r}

library(microbenchmark)

microbenchmark(
  "R" = reg1D(y, x),
  "Rccp" = parallelReg1D(y, x)
)

```